---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


## Loading and preprocessing the data

Load required libraries.

```{r}
library(ggplot2)
library(lattice)
```

Load the data into a simple data frame.

```{r}
unzip('activity.zip', overwrite=TRUE)
activity <- read.csv('activity.csv', stringsAsFactors=FALSE)
```

## What is mean total number of steps taken per day?

##### Calculate the total number of steps taken per day

```{r}
dates.factor <- as.factor(activity$date)
stepsPerDay <- tapply(activity$steps, activity$date, sum)
stepsPerDay <- data.frame(steps=stepsPerDay, date=levels(dates.factor))
summary(stepsPerDay, digits=12)
```

##### Make a histogram of the total number of steps taken each day

```{r}
g <- ggplot(stepsPerDay, aes(x=steps)) +
    geom_histogram(color='grey',fill='blue', alpha=0.50, binwidth=5000) +
    ggtitle('Histogram of Total Number of Steps per Day, ignoring missing values')
g
```

##### Calculate and report the mean and median of the total number of steps taken per day

```{r}
mean(stepsPerDay$steps, na.rm=TRUE)
median(stepsPerDay$steps, na.rm=TRUE)
```

## What is the average daily activity pattern?

Pack the calculation in a function, since we will need it again later.

```{r}
time.series <- function(activity) {
    dates.factor <- as.factor(activity$date)
    ndays <- nlevels(dates.factor)
    stepsPer5MinInterval <- tapply(activity$steps, activity$interval, sum, na.rm=TRUE)/ndays
    intervals <- sort(unique(activity$interval))
    stepsPer5MinInterval <- data.frame(steps=stepsPer5MinInterval, interval=intervals)
}
```

##### Calculate the average number of steps taken during each 5-minute interval, averaged across all days

```{r}
stepsPer5MinInterval <- time.series(activity)
```

##### Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
with(stepsPer5MinInterval, plot(interval, steps, type='l'))
title('Number of steps per 5-minute interal, averaged accross all days')
```

##### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
interval.max <- with(stepsPer5MinInterval, interval[which.max(steps)])
interval.max
```

## Imputing missing values

##### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with `NA`s)

As can be seen in the summary only the `steps` column has `NA` values.

```{r}
summary(activity)
na.steps <- sum(is.na(activity$steps))
```

There are `r na.steps` intervals where the step count is unavailable.


##### Create a new dataset that is equal to the original dataset but with the missing data filled in.

Replace all `NA`s in the `steps` column with the step mean for that 5-minute interval.

```{r}
# what activity observations have no step count available?
is.na.steps <- is.na(activity$steps)
# to which intervals do these NA's correspond?
intervalsForNASteps <- activity$interval[which(is.na.steps)]
# what indices have these intervals in the stepsPer5MinInterval data frame?
steps5MinIdx <- sapply(intervalsForNASteps, function(i) which(stepsPer5MinInterval$interval == i))
# set unavailable steps to the daily step average in the corresponding interval
activity$steps[is.na.steps] <- stepsPer5MinInterval$steps[steps5MinIdx]

summary(activity)
```

##### Make a histogram of the total number of steps taken each day

```{r}
dates.factor <- as.factor(activity$date)
stepsPerDay <- tapply(activity$steps, activity$date, sum)
stepsPerDay <- data.frame(steps=stepsPerDay, date=levels(dates.factor))
summary(stepsPerDay, digits=12)
```

```{r}
g <- ggplot(stepsPerDay, aes(x=steps)) +
    geom_histogram(color='grey',fill='blue', alpha=0.50, binwidth=5000) +
    ggtitle('Histogram of Total Number of Steps per Day,\nmissing values replaced with averages')
g
```

##### Calculate and report the mean and median total number of steps taken per day.

```{r}
mean(stepsPerDay$steps, na.rm=FALSE)
median(stepsPerDay$steps, na.rm=FALSE)
```

##### Do these values differ from the estimates from the first part of the assignment?

##### What is the impact of imputing missing data on the estimates of the total daily number of steps?

## Are there differences in activity patterns between weekdays and weekends?

##### Create a new factor variable in the dataset with two levels -- `weekday` and `weekend` indicating whether a given date is a weekday or weekend day.

```{r}
# day of week (Mo..Su)
dow <- as.POSIXlt(activity$date)$wday
# for each activity observation, takes it place during weekend?
weekend <- dow == 0 | dow == 6
# add factor
activity$daytype <- factor(weekend)
# change factor level names
levels(activity$daytype) <- c('weekday','weekend')
```

##### Calculate the average number of steps taken during each 5-minute interval, averaged across all weekend days and non-weekend days.

```{r}
wendStepsPer5MinInterval <- time.series(subset(activity, daytype == 'weekend'))
wendStepsPer5MinInterval$daytype = 'weekend'
wdayStepsPer5MinInterval <- time.series(subset(activity, daytype == 'weekday'))
wdayStepsPer5MinInterval$daytype = 'weekday'
stepsPer5MinInterval <- rbind(wendStepsPer5MinInterval, wdayStepsPer5MinInterval)
```

##### Make a panel plot containing a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r}
xyplot(steps ~ interval | daytype, data=stepsPer5MinInterval, type='l', layout=c(1,2))
```

